---
title: "BOO 2025 - Example Analysis"
subtitle: "Script 3: Sample QC - Answers"
date: "`r Sys.Date()`" 
author: 
  Adam Seuren
output:
  html_document:
    code_download: true
    theme: united
    highlight: tango
    df_print: paged
    toc: true
    number_sections: true
    toc_float:
      collapsed: false
      smooth_scroll: true
---

***

> In this script, you will perform sample-level quality control (QC), removing any poor quality samples and ensuring that experimental replicates are comparable to one another. 

***

# Setup

## Clean

As before, we perform several setup steps at the start of the script to ensure our work is reproducible and clear. 

**Exercise 1: Download the R markdown template and clean your environment:**

```{r clean}
rm(list=ls())
```

***

## Set variables

**Exercise 2: Create the following objects in your R environment:**

* `root_dir` - project folder
* `count_path` - location of the `countData` object within the project folder
* `cpm_path` - location of the `cpmData` object
* `metadata_path` - location of the `metaData` object

* `count_store` - location to save the `countData` object within the project folder
* `cpm_store` - location to save the `cpmData` object
* `metadata_store` - location to save the filtered `metaData` after QC

* `count_threshold` - minimum library size required for samples (600,000; 20% of the target sequencing depth)
* `corr_threshold` - required correlation between replicates (0.9)

```{r set-variables}
root_dir <- "/data/sinkelj1/BOO/Group_F/" 

count_path <- "countData.Rdata" 
cpm_path <- "cpmData.Rdata" 
metadata_path <- "metaData.Rdata"

count_store <- "countData.Rdata"
cpm_store <- "cpmData.Rdata"
metadata_store <- "metaData.Rdata" 

# Set count threshold to 600,000
count_threshold <- 6E5

# Set correlation threshold to 0.9
corr_threshold <- 0.9 
```

***

## Packages

Here, we load `tidyverse` and also a new package:

* `ggrepel` allows us labels in plots to "repel" each other and make visualizations clearer

**Exercise 3: Load `tidyverse` and `ggrepel` into your environment:**

```{r load-packages, warning=F, message=F}
library(tidyverse)
library(ggrepel)
```

***

## Load data

**Exercise 4: Load the count data, CPM data, and metadata into your environment:**

<details>
  <summary><strong>Hint</strong></summary>

  Make sure these are the ones your saved at the end of the probe QC.

</details>

```{r load-data, warning=FALSE, message=FALSE}
load(paste0(root_dir, count_path))
load(paste0(root_dir, cpm_path))
load(paste0(root_dir, metadata_path))
```

***

# Library size

## Check

Before applying any filters, it is good to perform some checks.

**Exercise 5: Check that the column names of `countData` match the `sample_ID` order in `metaData`:**

```{r order-check}
table(metaData$sample_ID == colnames(countData))
```

***

## Calculate

**Exercise 6: Now that we have removed unreliable and lowly expressed probes from `countData`, recalculate and save a new `lib_size` in the metadata:**

```{r calculate-lib}
summary(metaData$lib_size)

# Recalculate library size for each sample
metaData$lib_size <- colSums(countData)

summary(metaData$lib_size)
```

As expected, the library sizes are comparable (as most removed probes were lowly expressed), but slightly smaller than before.

***

## Distribution

**Exercise 7: Make a histogram of `lib_size`. What range of values does this variable take and is this reasonable for a TempO-Seq experiment?**

```{r lib-histogram}
# Summarize library size variable
summary(metaData$lib_size)

metaData %>% 
  # Create a histogram of library size
  ggplot(aes(x = lib_size)) +
  geom_histogram(fill = "lightblue", color = "black", bins = 30) +
  ggtitle("Histogram of library size values") + 
  xlab("Library size") +
  theme_bw()
```

The library size represents how many sequencing reads (or transcript fragments) in total were successfully generated and aligned to the TempO-Seq probe panel for that sample, irrespective of which gene they are assigned to. 

Library size in these samples ranges from 604,590 to 5,665,921 reads. An average sample has just over 3M reads. This is definitely reasonable for a TempO-Seq experiment, as the assay is highly targeted and efficient.

***

## Flag

Samples whose library size is below 20% of the targeted sequencing depth (`corr_threshold`; 600,000) should be flagged as having low reads.

**Exercise 8: Create a flag in `metaData` describing if samples have low reads not:**

```{r lib-flag}
metaData <- metaData %>% 
  # If library size is under threshold set TRUE, otherwise FALSE
  mutate(flagLowReads = ifelse(lib_size <= count_threshold, T, F))

# Tabulate the low read flag
table(metaData$flagLowReads)
```

In this case, all samples have sufficient reads and are kept in the data set. 

***

## Plot

It is good to visualize the library size for each sample, grouped by compound ID. This shows us whether samples are well above the threshold, and allows us to inspect the data more carefully.

**Exercise 9: Create a boxplot for the library sizes of each compound (including DMSO) and describe any patterns you identify:**

<details>
  <summary><strong>Hint</strong></summary>

  You can colour your boxplots by concentration to visualize patterns more clearly.

</details>

```{r lib-boxplot}
metaData %>%   
  # Ensure that concentration ID is treated as a factor
  mutate(conc_ID = factor(conc_ID)) %>%   
  # Plot compounds on the x axis and library size on the y axis
  ggplot(aes(x=compound_ID, y=lib_size)) + 
  # Create boxplots of the library size distributions
  geom_boxplot(aes(color=conc_ID), width=0.8) +  
  # Create a dashed line to represent the library size threshold
  geom_hline(aes(yintercept=count_threshold), color="grey5", linetype="dashed") +
  # Label any values below the threshold
  geom_text_repel(aes(x = compound_ID, y = lib_size, color = conc_ID),   
                   label=ifelse(metaData$lib_size < count_threshold, 
                                metaData$rep, "")) +
  # Set axis labels and title
  xlab("") + ylab("Library size") + ggtitle("Library size distributions") +    
  theme_bw()                                  
```

The highest 1,4-Benzenediol concentration shows reduced library sizes, with one sample nearing the threshold. While all samples passed the filter, the lower library sizes may reflect cytotoxicity, as dying cells yield less RNA and fewer viable cells are available for extraction under cytotoxic conditions.

***

# Replicate correlation

## log2CPM

The replicate correlation filter aims to remove any outlying replicates, with maximum pairwise correlations below the `corr_threshold` (set to 0.9). We usually perform this correlation analysis on the log2CPM values to ensure highly expressed genes do not have undue influence on the correlation values. A value of 1 is added to the CPM, to prevent issues arising from the fact that `log2(0)` is `-Inf`. 

**Exercise 10: Calculate and store the log2(CPM + 1) values in a `logcpmData` object:**

```{r log2cpm}
# Calculate log2(CPM + 1)
logcpmData <- log2(cpmData + 1)
```

***

## Pairwise correlations

In order to calculate pairwise correlations, each sample needs to be compared to the other replicates in its experimental group. We can do this by looping through `mean_ID`.

**Exercise 11: Calculate the pairwise replicate correlations for this data:**

<details>
  <summary><strong>Hint</strong></summary>

  The correlation can be calculated using `cor(cpmDataReps[,j], cpmDataReps[,k])` within an appropriate loop.

</details>

```{r pairwise-corr}
# Initialize the replicate filter output as a data frame
replicateFilterOutput <- data.frame()

# For each mean ID (experimental condition)
for(i in unique(metaData$mean_ID)){
  # Subset the meta data to keep only samples from this experiment
  metaDataReps <- metaData %>% 
    filter(mean_ID == i)
  
  # Subset the log2 CPM values to keep only these samples
  cpmDataReps <- logcpmData[, metaDataReps$sample_ID] 
  
  # Loop through each column in the CPM data  
  for(j in 1:ncol(cpmDataReps)){
    # In a pairwise fashion
    for(k in 1:ncol(cpmDataReps)){
      # Save the position in the loops that you are in
      sample_A <- colnames(cpmDataReps)[j]
      sample_B <- colnames(cpmDataReps)[k]
      
      # Don't calculate pairwise correlations between identical samples (since it will be 1)  
      if(sample_A != sample_B){
        # Calculate pairwise correlation values
        r2 <- cor(cpmDataReps[,j], cpmDataReps[,k])
        
        # Update the filter output data frame
        replicateFilterOutput <- rbind(
          replicateFilterOutput, 
          data.frame(mean_ID = i, 
                     sample_A = sample_A,
                     sample_B = sample_B,
                     r2 = r2))
      }
    }
  }
}

# Inspect output
head(replicateFilterOutput)
```

***

## Maximum

Each sample is judged by the best pairwise correlation it can achieve. If this is below `corr_threshold`, the sample should be flagged.

**Exercise 12: Calculate the `max_r2` for each sample and add it to the `replicateFilterOutput`:**

```{r max-r2}
replicateFilterOutput <- replicateFilterOutput %>% 
  # Separate sample name into compound and concentration using the underscore
  separate(sample_A, 
           into = c("Compound", "Conc_ID", NA, NA), 
           remove = F, 
           sep = "_") %>% 
  # If the compound is DMSO then keep only the first 5 letters (else we get DMSOHigh1 and DMSOHigh2 :D)
  mutate(Compound = ifelse(grepl("DMSO", Compound), substr(Compound,1,5), Compound)) %>% 
  # Group by sample
  group_by(sample_A) %>%
  # Save the maximum pairwise correlation for that sample
  mutate(max_r2 = max(r2, na.rm = T)) %>% 
  ungroup()

# Inspect output
summary(replicateFilterOutput$max_r2)
```

***

## Plot

**Exercise 13: Visualize the pairwise replicate correlations for each experimental conditions. Describe what you observe:**

```{r corr-boxplot}
replicateFilterOutput %>% 
  # Plot the sample ID against the pairwise correlation
  ggplot(aes(x = sample_A, y = r2)) +
  # Draw a boxplot of the pairwise correlation distribution
  geom_boxplot(color = "grey80") +
  # With points for the actual values
  geom_point(color = "grey60", size = 0.5) +
  # Highlight the maximum pairwise correlation for each sample
  geom_point(aes(y = max_r2, color = Conc_ID), 
             size = 1.5) +
  # Draw a line for the filter threshold
  geom_hline(aes(yintercept = corr_threshold), 
             color = "grey60", linetype = "dashed") +
  ylab("") + xlab("Sample ID") + ggtitle("Replicate correlations") +
  theme_bw() +
  # Do not print sample names
  theme(axis.text.x = element_blank()) +
  # Make a different plot for each compound, allowing the x-axis to change for different samples
  facet_wrap(~Compound, scales='free_x', nrow=2)
```

There is one replicate close to the threshold for the lowest concentration of 1,4-Benzenediol. Additionally, the highest concentration where there were signs of cytotoxicity also has low replicate correlations overall. This experimental condition is unlikely to be suitable for downstream analysis.

Looking at triethylene glycol and the DMSO controls, replicate correlation looks good and all samples can be kept in the data for analysis.

***

## Flag

**Exercise 14: Flag any samples that did not pass the replicate correlation filter in the `metaData`:**

<details>
  <summary><strong>Hint</strong></summary>

  You can merge the replicate correlation filter output with the metaData to create a `max_r2` column after some processing.

</details>

```{r corr-flag}
# Make a data frame of sample IDs and max r2
replicateFilterMerge <- replicateFilterOutput %>% 
  select(sample_ID = sample_A, max_r2) %>% 
  distinct()

# Merge with meta data
metaData <- left_join(metaData, replicateFilterMerge, 
                      by = "sample_ID") %>% 
  mutate(flagLowCorr = ifelse(max_r2 <= corr_threshold, T, F))

table(metaData$flagLowCorr)

# Inspect the flagged samples
metaData %>% 
  filter(flagLowCorr)
```

***

# Advanced questions

If you would like a bit more of a challenge, here are a few extra questions relating to the two sample QC steps above. However, you can also skip these, save your data, and move on to the PCA.

## Library size

**Exercise 14: What are the benefits of a sample having a higher library size and does this benefit apply to some genes more than others?**

A higher library size means greater average read depth, increasing the number of reads mapped to each gene. This improves statistical power and precision when estimating expression differences, particularly for moderately to highly expressed genes. It also enhances our ability to detect lowly expressed genes.

However, gene expression is highly skewed with some genes having thousands of reads and others only a few. Therefore, average depth (or total library size) alone doesn’t capture this variability. Examining the full distribution of read counts in a sample provides a more complete picture of sequencing depth and data quality:

```{r read-depth}
# Look only at one sample and select the ID and library size
metaData %>% 
  filter(sample_ID == "CS1101_C1_P1_R1") %>% 
  select(sample_ID, lib_size) %>% 
  # Calculate the average read depth for that sample
  mutate(read_depth = lib_size / nrow(countData),
         log10read_depth = log10(read_depth))

countData %>% 
  rownames_to_column(var="Probe") %>% 
  # Calculate log10 (count + 1) values for this sample
  mutate(CS1101_C1_P1_R1_log10 = log10(CS1101_C1_P1_R1 + 1)) %>% 
  ggplot() +
  # Plot a histogram
  geom_histogram(aes(x = CS1101_C1_P1_R1_log10),
                 fill = "lightblue", color = "black", bins = 30) +
  # Draw a line at the average read depth (on a log10 scale)
  geom_vline(aes(xintercept = 2.37), 
             color='red', linetype='dashed') +
  # Annotate the line with text
  annotate("text", x=3.1, y=1000, color='red',
           label="Average read depth: 233.1") +
  # Set the title, labels, and theme
  ggtitle("Distribution of counts across genes for one sample") +
  ylab("") + xlab("log10(counts + 1)") +
  theme_bw()
```

***

## Replicate correlation

Instead of looking at pairwise correlations, another way of measuring how good a replicate is is by comparing it to the average for that experimental condition. 

**Exercise 15: Calculate replicate correlation in this way and see if it alters the results of this filter. What is one benefit and downside of assessing replicate correlation in this manner?**

First, we calculate the correlation between each sample and its experimental mean.

```{r mean-corr}
# Initialize the replicate filter output as a data frame
replicateFilterOutput <- data.frame()

# Loop through experimental conditions
for(i in unique(metaData$mean_ID)){
  # Subset meta data
  metaDataReps <- metaData %>% filter(mean_ID == i)
  
  # Subset CPM data
  cpmDataReps <- logcpmData[, metaDataReps$sample_ID] 
    
  # Calculate the mean CPM for each probe in this experiment
  cpmDataReps$meanCPM = rowSums(cpmDataReps) / ncol(cpmDataReps)
  
  # Save the correlation between each sample and this average value
  corReps <- as.data.frame(cor(cpmDataReps)) %>% 
             select(repCor=meanCPM) %>% 
             rownames_to_column(var="sample_ID") %>% 
             filter(sample_ID != "meanCPM")
  
  # Update the output
  replicateFilterOutput <- rbind(replicateFilterOutput, corReps)
}

# Inspect
replicateFilterOutput
```

Then we can plot the output

```{r mean-plot}
replicateFilterOutput %>% 
  separate(sample_ID, 
           sep = "_", 
           into = c("Compound", "Conc_ID", "Plate", "Rep"), 
           remove=F) %>% 
  mutate(Conc_ID = factor(Conc_ID),
         Compound = ifelse(grepl("DMSO", Compound), substr(Compound,1,5), Compound)) %>% 
  ggplot(aes(x = Compound, y = repCor)) +
  geom_boxplot(aes(color = Conc_ID), width = 0.8) +
  geom_hline(aes(yintercept = corr_threshold), 
             color = "grey50", linetype = "dashed") +
  xlab("") + ylab("Correlation between replicates") +
  theme_bw()
```

Although assessing replicate correlation in this way yields similar results with lower correlations for the top and bottom concentration of 1,4-Benzenediol, in this instance all replicates now pass the replicate correlation filter. This demonstrates the somewhat arbitrary nature of these thresholds in practice.

A benefit of looking at correlation between each replicate and the mean is that there is a single value for each sample and no need to select a representative one (e.g. maximum pairwise correlation). However, the downside is that bad replicates can "drag down" the values of the fictitious average sample, especially for conditions where there are only a few replicates.

***

# Save

## Subset

**Exercise 16: Remove samples that did not pass the sample QC steps from your data:**

<details>
  <summary><strong>Hint</strong></summary>

  Don't forget to also subset the count and CPM data.

</details>

```{r any-flag}
# Subset the metadata to keep only high quality samples
metaData <- metaData %>% 
  filter(!flagLowReads & !flagLowCorr)

# Subset the count and CPM data
cpmData <- cpmData[ , metaData$sample_ID]
countData <- countData[ , metaData$sample_ID]

# Check dimensions
dim(metaData)
dim(countData)
dim(cpmData)
```

***

## Save

**Exercise 17: Save the updated data:**

```{r save-metadata}
# Save
save(metaData, file=paste0(root_dir, metadata_store))
save(cpmData, file=paste0(root_dir, cpm_store))
save(countData, file=paste0(root_dir, count_store))
```

***

# Session Info

**Exercise 18: Print your session info at the end of the script, knit the R markdown document, and push it to GitHub:**

```{r session-info}
# Session info
devtools::session_info()
```

***

That is the end of the Sample QC. Example answers will be available from the `BOO_template` GitHub on Tuesday. 

Next, please move on to the PCA using `04_PCA_Outline.Rmd`.

***

